---
title: "Describing Data"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r, echo = FALSE, message = FALSE}
library(ggplot2)
library(Hmisc)
```

How do you summarize data? What do you look for when investigating data?

# Types of Data

## Statistics

### quantity
* discrete - count
* continuous - measure

### quality
* binary - true/false
* nominal - unordered
* ordinal - ordered

These are categorized into discrete variables ("factor" in R).

## Programming

* character - surround with quotes (' or ")
* numeric
* logical - TRUE or FALSE

For simplicity, I'm ignoring "integer" and "complex".

### unquantifiable
* infinite - Inf
* missing - NA
* undefined - NULL
* not a number - NaN (Inf - Inf)

# Importing Data

## Raw Data
* delimited (comma, tab)

## Binary Data
* applications (Excel, statistical software, database)

### Examples
* scan
* read.table
* read.csv
* readxl::read_excel
* load

# Exporting Data

## Examples
* write.csv
* save

```{r}
# read a CSV file from URL
bat <- read.csv('https://github.com/vubiostat/softball-stats/raw/csv/team_batting_stats_season.csv')
# produce a random temporary filename
tf <- tempfile()
# write data set to temporary file; don't include row names
write.csv(bat, file = tf, row.names = FALSE)
# `read.csv` is more appropriate, but we can read in one line (\n) at a time with `scan`
scan(tf, what = '', n = 6, sep = '\n')
# show the first few rows of our data set (default is 6 rows)
head(bat)
```

# Describing Data

## Descriptive statistics

The summary of continuous data usually includes mean, standard deviation, and quantiles. It may also include number of observations and number of observations missing. Discrete data is often summarized with a frequency table.

## Table 1

In a journal article the first table of data typically presents descriptive statistics of variables grouped by the exposure variable.

### Examples

For the next few examples, we'll download SAS data sets from the CDC's NHANES program. We'll take a look at participants' age and gender (demographics), as well as their standard biochemistry profile (drawn blood lab samples).

SAS is a proprietary (not open source nor free) alternative to R.

Here's a link to the description of the profile data: [biopro](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_BIOPRO.htm)

```{r}
# use `haven` package and `read_xpt` function to read SAS file
# restrict to three columns: userid (SEQN), gender (RIAGENDR), age (RIDAGEYR)
# use `as.data.frame` to convert from `tibble` to `data.frame`
demo <- as.data.frame(haven::read_xpt('https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_DEMO.xpt'))[, c('SEQN','RIAGENDR','RIDAGEYR')]
# read profile data
dat <- as.data.frame(haven::read_xpt('https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_BIOPRO.xpt'))
# merge two data sets by common userid column (SEQN)
profile <- merge(demo, dat)
```

```{r}
# use `Hmisc` package and `label` function to provide full labels for variables in our data set
# this is tedious unless the labels can be pulled from somewhere
profile <- within(profile, {
  label(SEQN)     <- 'Respondent Sequence Number'
  label(RIAGENDR) <- 'Gender'
  label(RIDAGEYR) <- 'Age in years at screening'
  label(LBXSATSI) <- 'Alanine Aminotransferase (ALT) (IU/L)'
  label(LBDSATLC) <- 'ALT Comment Code'
  label(LBXSAL)   <- 'Albumin, refrigerated serum (g/dL)'
  label(LBDSALSI) <- 'Albumin, refrigerated serum (g/L)'
  label(LBXSAPSI) <- 'Alkaline Phosphatase (ALP) (IU/L)'
  label(LBXSASSI) <- 'Aspartate Aminotransferase (AST) (IU/L)'
  label(LBXSC3SI) <- 'Bicarbonate (mmol/L)'
  label(LBXSBU)   <- 'Blood Urea Nitrogen (mg/dL)'
  label(LBDSBUSI) <- 'Blood Urea Nitrogen (mmol/L)'
  label(LBXSCLSI) <- 'Chloride (mmol/L)'
  label(LBXSCK)   <- 'Creatine Phosphokinase (CPK) (IU/L)'
  label(LBXSCR)   <- 'Creatinine, refrigerated serum (mg/dL)'
  label(LBDSCRSI) <- 'Creatinine, refrigerated serum (umol/L)'
  label(LBXSGB)   <- 'Globulin (g/dL)'
  label(LBDSGBSI) <- 'Globulin (g/L)'
  label(LBXSGL)   <- 'Glucose, refrigerated serum (mg/dL)'
  label(LBDSGLSI) <- 'Glucose, refrigerated serum (mmol/L)'
  label(LBXSGTSI) <- 'Gamma Glutamyl Transferase (GGT) (IU/L)'
  label(LBDSGTLC) <- 'GGT Comment Code'
  label(LBXSIR)   <- 'Iron, refrigerated serum (ug/dL)'
  label(LBDSIRSI) <- 'Iron, refrigerated serum (umol/L)'
  label(LBXSLDSI) <- 'Lactate Dehydrogenase (LDH) (IU/L)'
  label(LBXSOSSI) <- 'Osmolality (mmol/Kg)'
  label(LBXSPH)   <- 'Phosphorus (mg/dL)'
  label(LBDSPHSI) <- 'Phosphorus (mmol/L)'
  label(LBXSKSI)  <- 'Potassium (mmol/L)'
  label(LBXSNASI) <- 'Sodium (mmol/L)'
  label(LBXSTB)   <- 'Total Bilirubin (mg/dL)'
  label(LBDSTBSI) <- 'Total Bilirubin (umol/L)'
  label(LBDSTBLC) <- 'Total Bilirubin Comment Code'
  label(LBXSCA)   <- 'Total Calcium (mg/dL)'
  label(LBDSCASI) <- 'Total Calcium (mmol/L)'
  label(LBXSCH)   <- 'Cholesterol, refrigerated serum (mg/dL)'
  label(LBDSCHSI) <- 'Cholesterol, refrigerated serum (mmol/L)'
  label(LBXSTP)   <- 'Total Protein (g/dL)'
  label(LBDSTPSI) <- 'Total Protein (g/L)'
  label(LBXSTR)   <- 'Triglycerides, refrig serum (mg/dL)'
  label(LBDSTRSI) <- 'Triglycerides, refrig serum (mmol/L)'
  label(LBXSUA)   <- 'Uric acid (mg/dL)'
  label(LBDSUASI) <- 'Uric acid (umol/L)'
})
```

We'll remove some of the redundant columns (something we would not typically do).

```{r}
# specify column names
col2rm <- c(
'LBDSATLC','LBDSALSI','LBDSBUSI','LBDSCRSI',
'LBDSGBSI','LBDSGLSI','LBDSGTLC','LBDSIRSI',
'LBDSPHSI','LBDSTBSI','LBDSTBLC','LBDSCASI',
'LBDSCHSI','LBDSTPSI','LBDSTRSI','LBDSUASI'
)
# `setdiff` compares two vectors (A & B) and returns items in vector A not found in B
# which in this case means find the columns not present in `col2rm`
# with the end result being those columns (in `col2rm`) are removed from our data set
profile <- profile[,setdiff(names(profile), col2rm)]
```

After adding labels and removing several columns, we're ready to take a closer look at our data. We'll use the `Hmisc` package to show descriptive statistics and provide table 1 output. There are many packages that could provide this functionality.

```{r}
# to format our output specifically for web-view, we can wrap it with `html`
html(describe(profile))
```

```{r}
# we use `setdiff` to remove SEQN and RIAGENDR from our vector
# if the vector was c('A','B','C'), our paste would produce the string "A + B + C"
cats <- paste(setdiff(names(profile), c('SEQN','RIAGENDR')), collapse = ' + ')
# we now paste "~ RIAGENDR" to the end of our string
# the "~" has a special meaning, where everything on the left are dependent variables
# the variable on the right (RIAGENDR) is our exposure variable
# we will use it to split (stratify) our participants by gender
form <- sprintf('%s ~ RIAGENDR', cats)
# `summaryM` produce our table 1
m <- summaryM(as.formula(form), data = profile)
# m <- summaryM(as.formula(form), data = profile, overall = TRUE)
# tableone::CreateTableOne(vars = names(profile), strata = c('RIAGENDR'), data = profile)
```

```{r}
# we use some additional arguments to customize our output
# "prmsd" will print the mean and standard deviation
# "prn" and "prN" are both related to including the number of non-missing observations
html(m, round = 2, prmsd = TRUE, prn = FALSE, prN = TRUE)
```

# Visualizing Data

There are many ways we can produce a plot to visually inspect our data. Here's one example, where we take a closer look at the bicarbonate variable. We'll use the `ggplot2` package to create a density plot.

```{r}
ggplot(data = profile, aes(x = LBXSC3SI)) + geom_density()
```
