---
title: "Visualizing Data - ggplot2"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r, echo = FALSE, message = FALSE}
library(ggplot2)
```

In ggplot, the values in a column of a data.frame will be the "ink" in the plot.

Do we want to see raw data, a summary of data, or a combination of both?

* include raw data whenever possible (i.e., not millions of points)

Who is the audience, and is the plot interpretable to them?

Do we want to visualize a single variable, the relationship of two variables, or the relationship of more than two variables?

X and Y-axis

* the x-axis is horizontal (east-west)
* the y-axis is vertical (north-south)
* if we want to show that one variable influences the other, we typically will put the variable of influence (independent variable) on the x-axis and the variable being influenced (dependent variable) on the y-axis

## Components of ggplot

[ggplot2 cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)

* Geoms
* Aesthetics
* Stats
* Scales
* Faceting
* Themes

## Major function

* ggplot

## Common Visualizations

In ggplot, the type of plot references a "geom".

geom | variable 1 | variable 2
-----|------------|-----------
density | continuous | NA
histogram | continuous | NA
bar | discrete | NA
point | continuous | continuous
boxplot | discrete | continuous
dotplot | discrete | continuous
violin | discrete | continuous

### Data

Data on 671 infants with very low (<1600 grams) birth weight from 1981-87 were collected at Duke University Medical Center.

```{r}
vlbw <- read.csv('https://github.com/couthcommander/Bios6301/raw/main/datasets/vlbw.csv', row.names = 1)
head(vlbw)
```

### Examples

For the given plots, are we showing the raw data or a summary of the data? If it's a summary, what type of summary is being shown?

```{r}
ggplot(data = vlbw) + geom_density(mapping = aes(x = bwt))
```

```{r}
ggplot(data = vlbw) + geom_histogram(mapping = aes(x = bwt))
```

```{r}
ggplot(data = vlbw) + geom_bar(mapping = aes(x = race))
```

```{r}
ggplot(data = vlbw) + geom_point(mapping = aes(x = gest, y = bwt))
```

```{r}
ggplot(data = vlbw) + geom_boxplot(mapping = aes(x = sex, y = bwt))
```

```{r}
ggplot(data = vlbw) + geom_dotplot(mapping = aes(x = sex, y = bwt), binaxis = 'y', binwidth = 10)
```

```{r}
ggplot(data = vlbw) + geom_violin(mapping = aes(x = sex, y = bwt))
```

We can also add lines, shapes, and text.

### Overlapping with `jitter`

We can provide overlapping geometries. Order matters.

```{r}
ggplot(data = vlbw, aes(x = sex, y = bwt)) + geom_jitter() + geom_boxplot()
```

```{r}
ggplot(data = vlbw, aes(x = sex, y = bwt)) + geom_boxplot() + geom_jitter()
```

### Aesthetics

We can use aesthetics to map variables to "x" and "y" as well as color, size, and shape.

```{r}
ggplot(data = vlbw) + geom_point(mapping = aes(x = gest, y = bwt, color = ivh))
```

```{r}
ggplot(data = vlbw) + geom_point(mapping = aes(x = gest, y = bwt, color = sex, shape = sex))
```

```{r}
ggplot(data = vlbw) + geom_point(mapping = aes(x = gest, y = bwt, size = lol))
```

### Faceting

If we want to compare plots by a categorical variable, we can use faceting.

Faceting introduces the "~" (tilde) character. Think of it as having a left-hand side (LHS) and right-hand side (RHS).

```{r}
bestRows <- complete.cases(vlbw[,c('gest','bwt','sex')])
p <- ggplot(data=vlbw[bestRows,]) + geom_point(mapping = aes(x=gest, y=bwt))
p
```

A plot for each race.

```{r}
p + facet_wrap(~ race)
```

A plot for each race/sex combination.

```{r}
p + facet_wrap(~ race + sex)
```

A plot for each race/sex combination, organized into row (LHS, sex) and column (RHS, race).

```{r}
p + facet_grid(sex ~ race)
```

A plot for each race, organized into columns (RHS).

```{r}
p + facet_grid(. ~ race)
```

A plot for each race, organized into rows (LHS).

```{r}
p + facet_grid(race ~ .)
```

### Theme

```{r}
ggplot(data = vlbw) + geom_point(mapping = aes(x = bwt, y = pltct)) +
  labs(title="Platelet count as function of birth weight") +
  theme(axis.title.x = element_text(face = "bold", colour="red"), axis.title.y = element_text(size = 20, angle = 0))
```

### More advanced examples

```{r}
ggplot(data=vlbw[bestRows,]) + geom_point(mapping = aes(x=gest, y=bwt)) + geom_smooth(mapping = aes(x=gest, y=bwt))
```

Rather than applying a "geom", we could apply a "stat" instead. One useful plot is the "ECDF".

```{r}
ggplot(vlbw, aes(gest)) + stat_ecdf()
```

Or even add both a "geom" and "stat".

```{r}
ggplot(vlbw, aes(apg1, bwt)) + geom_point() + stat_summary(fun.data = "mean_sdl", geom = "errorbar", color = "red")
```
