---
title: "Visualizing Data - base graphics"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

In base graphics, you start with an empty canvas and add layers.

## Major functions

* plot

## Main arguments to "plot" function

* xlab - x-axis label
* ylab - y-axis label
* xlim - min and max values for x-axis
* ylim - min and max values for y-axis
* col - color
* pch - point char
* lty - line type
* lwd - line width

## Components (layers) of a plot in base graphics

* points
* lines
* abline
* segments
* rect
* arrows
* axis
* text
* mtext (margin text)
* legend

### Notes about the axis

While a plot has two axes (x and y), the plotting location for a given axis is referenced by a specific order and number in base graphics.

* axis 1, bottom
* axis 2, left
* axis 3, top
* axis 4, right

While you could use all four axis locations, generally you would only use 1 and 2.

## A few specific plot functions

* hist
* boxplot
* barplot

## Additional functions

* par - graphical parameters

### Examples

We'll revisit the low birth weight data. Let's create a scatterplot of gestational "age" by "birth weight".

```{r}
plot(vlbw[,'gest'], vlbw[,'bwt'])
```

Let's also include "sex" by changing the color and point character. We'll try to mimic the output we would get from ggplot.

```{r}
vlbw <- read.csv('https://github.com/couthcommander/Bios6301/raw/main/datasets/vlbw.csv', row.names = 1)
vlbw[,'race'] <- as.factor(vlbw[,'race'])
op <- par(mar = c(5,4,1,1) + 0.1)
plot(vlbw[,'gest'], vlbw[,'bwt'], xlab='', ylab='', axes=FALSE, type='n')
rect(20, 350, 40.75, 1625.5, col = '#E1E1E1')
points(vlbw[,'gest'], vlbw[,'bwt'],
    col = ifelse(vlbw[,'sex']=='male', "#00BFC4", "#F8766D"),
    pch = ifelse(vlbw[,'sex']=='male', 17, 16)
)
axis(1)
axis(2, at = c(400, 800, 1200, 1600))
mtext("Gestational Age", side=1, line=3)
mtext("Birth Weight", side=2, line=3)
lab <- c('Male','Female')
legend("bottomright", lab, pch = c(17,16), col= c('#00BFC4','#F8766D'))
box()
grid(9, 8, col = 'white')
par(op)
```

We can create a barplot by passing a variable into the `table` function first.

```{r}
barplot(table(vlbw[,'race']))
```

We can even take the `table` of two variables as the input for `barplot`.

```{r}
barplot(table(vlbw[,'sex'], vlbw[,'race']), beside = TRUE, legend.text = TRUE)
```

### Data and distributions

Let's examine a few built-in distributions in R that we can use to generate data. In these cases we'll want to see a histograph and boxplot.

I'll use the `layout` function to include several plots in the same output.

```{r}
n <- 10000
data_norm <- rnorm(n, mean = 0, sd = 1)
data_unif <- runif(n, min = 0, max = 1)
data_pois <- rpois(n, 4)

layout(matrix(1:6, 3, 2))
hist(data_norm)
boxplot(data_norm)

hist(data_unif)
boxplot(data_unif)

hist(data_pois)
boxplot(data_pois)
```

```{r}
hist(data_norm)
abline(v = -2, lty = 2, col = 'red')
abline(v = 2, lty = 2, col = 'red')
segments(-2, 1600, 2, 1600)
segments(-2, 1550, -2, 1650)
segments(2, 1550, 2, 1650)
text(3, 1600, '95% of data\nbetween\n-1.96 and 1.96')
quantile(data_norm, c(0.025, 0.975))
```
