---
title: "R programming"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Control structures

* conditional - if
* repeatable code chunk - function
* loops - for, while, *apply

## Example data

```{r}
# `expand.grid` will build combinations for us
x <- expand.grid(intervention = c("control", "treatment"), gender = c("M", "F"))
x[,'los'] <- c(5,3,9,6)
x[,'smoker'] <- c(0,1,1,0)
x
```

## An `if` statement

An "if" statement will run a chunk of code if a condition results in TRUE. We can test multiple statements with "and" (&&) and "or" (||). An "and" condition requires that both statements are TRUE. An "or" condition requires that at least one statement is TRUE.

```{r}
gender <- 'M'
smoker <- 1
if(gender == 'M' && smoker == 1) {
  grp <- 'male.smoker'
} else if(gender == 'M' && smoker != 1) {
  grp <- 'male.nonsmoker'
} else if(gender == 'F' && smoker == 1) {
  grp <- 'female.smoker'
} else if(gender == 'F' && smoker != 1) {
  grp <- 'female.nonsmoker'
} else {
  grp <- NA
}
grp
```

## A user-defined function

We can easily turn the above into a function.

```{r}
classifyGenderSmoker <- function(gender, smoker) {
  if(gender == 'M' && smoker == 1) {
    grp <- 'male.smoker'
  } else if(gender == 'M' && smoker != 1) {
    grp <- 'male.nonsmoker'
  } else if(gender == 'F' && smoker == 1) {
    grp <- 'female.smoker'
  } else if(gender == 'F' && smoker != 1) {
    grp <- 'female.nonsmoker'
  } else {
    grp <- NA
  }
  grp
}
classifyGenderSmoker('F', 0)
```

## A `for` loop

```{r}
# number of observations
n <- nrow(x)
# initialize character vector of proper length
grp <- character(n)
for(i in seq(n)) {
  # retrieve the "i"th row, and gender and smoker columns
  gender <- x[i,'gender']
  smoker <- x[i,'smoker']
  # save result as "i"th element of vector
  grp[i] <- classifyGenderSmoker(gender, smoker)
}
grp
```

## A `while` loop

```{r}
# find first spot with "smoker"
i <- 1
while(x[i,'smoker'] == 0) {
  i <- i + 1
}
i
```

## Apply functions

"apply" functions are convenience functions in that a single apply statement works like a pre-defined `for` loop running a specified function.

```{r}
apply(x[,c('los','smoker')], MARGIN = 2, FUN = sd)
```

In the above, why is MARGIN = 2?

```{r}
lapply(split(x[,'los'], x[,'intervention']), mean)
```

```{r}
sapply(split(x[,'los'], x[,'intervention']), mean)
```

```{r}
tapply(x[,'los'], x[,'intervention'], mean)
```

```{r}
mapply(classifyGenderSmoker, x[,'gender'], x[,'smoker'])
```

# Dates and times

## `Date` Class

Internally, `Date` stores time as an integer value, which represents the number of days since January 1, 1970

```{r}
unclass(Sys.Date())
as.Date("09/30/2012", format="%m/%d/%Y") # Custom formatting
as.Date(41180, origin="1900-01-01") # Choose a different origin
```

## Date-Time Data

* `POSIXct`, calendar time, stores time as the number of *seconds* since the origin
* `POSIXlt`, local time,  stores a list of time attributes, which can be indexed

```{r}
unclass(Sys.time())
unclass(as.POSIXlt(Sys.time()))
```

## Date-Time Formatting

See documentation on `strptime`

```{r}
as.POSIXct("080406 10:11", format = "%y%m%d %H:%M")
as.POSIXct("2008-04-06 10:11:01 PM", format = "%Y-%m-%d %I:%M:%S %p")
as.POSIXct("08/04/06 22:11:00", format = "%m/%d/%y %H:%M:%S")
```

```{r}
now <- Sys.time()
# create the last 7 days at this time
past <- now - 86400 * 1:7
difftime(now, past, units="secs")
difftime(now, past, units="hours")
difftime(now, past, units="days")
```

# Data processing

## Setting and extracting

Remember this? `gender == 'M' && smoker == 1` We can apply this conditional check to our data.frame.

```{r}
x[,'gender'] == 'M' && x[,'smoker'] == '1'
x[,'gender'] == 'M' & x[,'smoker'] == '1'
```

We can extract rows based on conditions.

```{r}
x[x[,'gender'] == 'M' & x[,'smoker'] == '1',]
```

As an aside, sometimes we encounter missing data (NA values). Let's add a new row and see what happens when we test for "gender".

```{r}
x1 <- rbind(x, data.frame(intervention = 'control', gender = NA, los = 6, smoker = 0))
# the new row results in NA
x1[,'gender'] == 'M'
# if we use this to extract, our result is incorrect
x1[x1[,'gender'] == 'M',]
# we need to modify our test to check for NA as well
x1[!is.na(x1[,'gender']) & x1[,'gender'] == 'M',]
```

What if we want to count the occurrences of a condition? Instead of "extracting", we can use the `sum` function.

```{r}
sum(!is.na(x1[,'gender']) & x1[,'gender'] == 'M')
# we can ignore the check for missing by adding the "na.rm" argument
sum(x1[,'gender'] == 'M', na.rm = TRUE)
```

We can also extract rows based on row number.

```{r}
# show the 3rd row first, then 4th, 2nd, 1st (i.e., change row order)
x[c(3,4,2,1),]
# show row numbers based on sorted order of "los" column
order(x[,'los'])
# show row numbers that satisfy condition
which(x[,'gender'] == 'F')
# order data by "los"
x[order(x[,'los']),]
# subset data by gender "F"
x[which(x[,'gender'] == 'F'),]
```

We can do the same with setting/assignment.

```{r}
x[x[,'gender'] == 'M' & x[,'smoker'] == '1','GenderSmokerGrp'] <- 'M.s'
x[x[,'gender'] == 'M' & x[,'smoker'] != '1','GenderSmokerGrp'] <- 'M.ns'
x[x[,'gender'] == 'F' & x[,'smoker'] == '1','GenderSmokerGrp'] <- 'F.s'
x[x[,'gender'] == 'F' & x[,'smoker'] != '1','GenderSmokerGrp'] <- 'F.ns'
x
```

Using clever logic, we can actually set our "GenderSmokerGrp" variable with a single statement.

```{r}
(x[,'gender'] == 'M') * 2 + x[,'smoker'] + 1
x[,'GenderSmokerGrpOth'] <- factor((x[,'gender'] == 'M') * 2 + x[,'smoker'] + 1, labels = c('F.ns', 'F.s', 'M.ns', 'M.s'))
x
```

## Merging two data.frames

To combine data frames based on the values of common variables, we can use the built-in `merge` function. By default, `merge` joins rows of the data frames based on the values of the columns that the data frames have in common.

```{r}
df1 <- data.frame(a=c(1,2,4,5,6),x=c(9,12,14,21,8))
df2 <- data.frame(a=c(1,3,4,6),y=c(8,14,19,2))
merge(df1, df2)
merge(df1, df2, all=TRUE)
merge(df1, df2, all.x=TRUE)
merge(df1, df2, all.y=TRUE)
```

## Finding string matches

If we want to find a string pattern within a vector (or column), we can use variations of the `grep` function depending on if we want the location, value, or conditional/logical.

When working with character data, it may also help to transform all strings to lower/upper-case with `tolower`/`toupper`.

```{r}
grep("ns", x[,'GenderSmokerGrp'])
grep("ns", x[,'GenderSmokerGrp'], value = TRUE)
grepl("ns", x[,'GenderSmokerGrp'])
tolower("This Is A String With Capitalization")
```

Another common function to use with matching patterns is the `gsub` function for string substitution.

```{r}
gsub("ns", "nons", x[,'GenderSmokerGrp'])
```

Both `grep` and `gsub` support regular expressions (view documentation of `?regex`).
