---
title: "Tests on mean values"
author: Shilin Zhao and Leena Choi
output:
  rmdformats::robobook:
    highlight: kate
    toc_depth: 3
    number_sections: no
    toc_float:
      collapsed: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  markdown:
    wrap: 72
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```
## Introduction

Hypothesis tests and p-values

-   Motivation

    -   Population mean and sample mean
    -   Understand the question
    -   Explore the data

-   Comparing **one-sample mean** to a known mean:

    -   One-sample t-test (parametric)
    -   One-sample Wilcoxon test (non-parametric)

-   Comparing the **means of two independent groups**

    -   Unpaired two samples t-test (parametric)
    -   Unpaired two-samples Wilcoxon test (non-parametric)

-   Comparing the **means of paired samples**

    -   Paired samples t-test (parametric)
    -   Paired samples Wilcoxon test (non-parametric)

-   Comparing the **means of more than two groups**

    -   One-way ANOVA test (parametric)

    -   Kruskal-Wallis test (non parametric)

## Data example 1

Studies of pulmonary function in adults who are long-time smokers have
shown that smoking tends to impair lung function. A question then arises
as to whether similar effects can be detected in children who smoke. We
will use **fev** data that include measures of lung function, forced
expiratory volume (FEV), which were collected from 654 children seen for
a routine checkup in a pediatric clinic. The study subjects were asked
whether they were current smokers. FEV measures the volume of air you
can blow out of your lungs.

```{r, message = FALSE}
library(Hmisc)
#fev=read.csv("./data/fev.csv",header=TRUE,row.names=1)
fev <- read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/fev.csv",header=TRUE,row.names=1)
label(fev$age) <- 'subject age at time of measurement (years)'
label(fev$fev) <- 'measured FEV (liters per second)'
label(fev$height) <- 'subject height at time of measurement (inches)'
label(fev$sex) <- 'sex (1 = male, 2 = female)'
label(fev$smoke) <- 'smoker (1 = smoker, 2 = non-smoker)'
head(fev)
html(describe(fev))

```

## Comparing one-sample mean to a known mean

**Question**: Is the mean value of age in this data different from 10?

### Check data distribution to decide which test to use

```{r}

hist(fev$age, main="", xlab="Age")
abline(v=10, lty=2, col="red")

```

### One-sample t-test (parametric)

```{r}
#T stastic caculation
sampleMean=mean(fev$age)
sampleSD=sd(fev$age)
sampleN=nrow(fev)
populationMean=10

TValue=(sampleMean-populationMean)/(sampleSD/sqrt(sampleN))
TValue

```

```{r}
#one sample t test
t.test(fev$age, mu=populationMean)

```

The output of t.test() function is a list containing the following
components:

-   statistic: the value of the t-statistic

-   p.value: the p-value for the test

-   conf.int: a confidence interval for the mean appropriate to the
    specified alternative hypothesis.

-   estimate: the estimated mean for one-sample test or difference in
    means for the two groups being compared (in the case of two-sample
    t-test).

### One-sample Wilcoxon test (non-parametric)

```{r}

wilcox.test(fev$age, mu = populationMean)

```

## Comparing the means of two independent groups

**Question**: Is the mean value of fev between smokers and non-smokers
different?

### Check data distribution to decide which test to use

```{r}

hist(fev$fev[which(fev$smoke==1)], main="Smokers", xlab="FEV")
hist(fev$fev[which(fev$smoke==2)], main="Non-smokers", xlab="FEV")
boxplot(fev$fev~fev$smoke, main="", ylab="FEV", xlab="Smoking behaviors")

tapply(fev$fev, fev$smoke,summary)

```

### Two-sample t-test (parametric)

T-statistic calculation

```{r}

#T statistic calculation
dataX=(fev$fev[which(fev$smoke==1)])
dataY=(fev$fev[which(fev$smoke==2)])

meanX=mean(dataX)
meanY=mean(dataY)
sdX=sd(dataX)
sdY=sd(dataY)
nX=length(dataX)
nY=length(dataY)

TValue=(meanX-meanY)/(sqrt(sdX^2/nX+sdY^2/nY))
TValue

```

Two-sample t-test

```{r}

t.test(dataX, dataY, alternative = "two.sided", paired = FALSE)

```

### Wilcoxon rank sum test (non-parametric):

Wilcoxon rank sum test is a non-parametric test for means of independent
two samples.

```{r}

wilcox.test(dataX,dataY)

```

## Data example 2

We will use the **sot-covid** data used in "01_introduction". The goal
of this study was to compare humoral and cellular immune responses to
two doses of the SARS-CoV-2 mRNA vaccine between solid organ transplant
(SOT) recipients and healthy controls (HCs). See "01_introduction" for
the details of this data.

```{r}

#load data
dataForAnalysis=read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/sot-covid.csv",header=TRUE,row.names=1)

head(dataForAnalysis)

```

## Comparing the mean of paired samples

Let's compare Immunoglobulin G (IgG) to SARS-CoV-2 spike
receptor-binding domain (RBD) values between the baseline (`rbd1`) and
three weeks following first vaccine dose (`rbd2`).

### Paired t-test (parametric)

Data summary for changes in RBD between the baseline and three weeks
following first vaccine dose

```{r}

summary(dataForAnalysis$rbd2-dataForAnalysis$rbd1)

hist(dataForAnalysis$rbd2-dataForAnalysis$rbd1, main="", xlab="Changes in RBD between the baseline and 3-weeks follow-up")

plot(dataForAnalysis$rbd1, dataForAnalysis$rbd2, main="", xlab="RBD at baseline", ylab="RBD at follow-up")
abline(0, 1, lty=2, col="red")

```

t-statistic calculation on "difference"

```{r,echo=TRUE}
#T stastic caculation
diffValue=dataForAnalysis$rbd1-dataForAnalysis$rbd2

cohortMean=mean(diffValue)
cohortSD=sd(diffValue)
cohortN=length(diffValue)
populationMean=0

TValue=(cohortMean-populationMean)/(cohortSD/sqrt(cohortN))
TValue

```

Paired t-test

```{r}

t.test(dataForAnalysis$rbd1, dataForAnalysis$rbd2, paired = TRUE)

```

You will get same results by one sample t-test to compare the difference
with mean=0

```{r}

t.test(dataForAnalysis$rbd1-dataForAnalysis$rbd2, mu=0)

```

### Wilcoxon signed rank test (non-parametric)

Wilcoxon signed rank test is a non-parametric test for the mean of
paired two samples.

```{r}

wilcox.test(dataForAnalysis$rbd1, dataForAnalysis$rbd2, paired = TRUE)


```

## Analysis of Variance (ANOVA): testing means for three or more samples

The ANOVA test assumes that the data are normally distributed and the
variance across groups are homogeneous.

Let's compare the baseline RBD values (`rbd1`) between different groups
of organ transplant patients, as well as three weeks following first
vaccine dose (`rbd2`).

***organ:*** Transplant organ type (0 = healthy control, 1 = kidney, 2 =
liver, 3 = heart, 4 = lung).

### Data visualization

```{r}
boxplot(rbd1 ~ organ, data = dataForAnalysis, main="", xlab="Transplant organ type", ylab="RBD at the baseline")

boxplot(rbd2 ~ organ, data = dataForAnalysis, main="", xlab="Transplant organ type", ylab="RBD at the 3-weeks follow-up")

boxplot((rbd2-rbd1) ~ organ, data = dataForAnalysis, main="", xlab="Transplant organ type", ylab="RBD changes between 3-weeks follow-up and baseline")

hist(dataForAnalysis$rbd1, main="", xlab="RBD at the baseline")
hist(dataForAnalysis$rbd2, main="", xlab="RBD at the 3-weeks follow-up")

```

### One-way ANOVA

```{r}

dataForAnalysis$organ= factor(dataForAnalysis$organ)

#difference of baseline rbd between organ types
anovaResult1 <- aov(rbd1 ~ organ, data = dataForAnalysis)
summary(anovaResult1)

#difference of rbd at three weeks following first vaccine dose between organ types
anovaResult2 <- aov(rbd2 ~ organ, data = dataForAnalysis)
summary(anovaResult2)

#difference of rbd change in 3-weeks follow-up and baseline between organ types
anovaResult3 <- aov((rbd2-rbd1) ~ organ, data = dataForAnalysis)
summary(anovaResult3)

```

### Tukey multiple pairwise-comparisons

```{r}

#TukeyHSD(anovaResult1)
TukeyHSD(anovaResult2)

```

### Kruskal-Wallis test

Kruskal-Wallis test is a non-parametric alternative to one-way ANOVA
test, which extends Wilcoxon rank sum test when there are more than two
groups.

```{r}
#baselilne rbd
kruskal.test(rbd1 ~ organ, data = dataForAnalysis)

#rbd at three weeks following first vaccine dose
kruskal.test(rbd2 ~ organ, data = dataForAnalysis)

```

```{r,warning=FALSE}

pairwise.wilcox.test(dataForAnalysis$rbd2 , dataForAnalysis$organ,
                 p.adjust.method = "BH")

```

## Summary

Data type: continuous data;

Normality: Parametric or non-parametric?

Groups: One, Two or more groups?

Paired or not paired?

**Interpretation and limitations:**

-   Sample Size and Power: Smaller sample sizes may lead to higher
    variability and less precise estimates.

-   Type I and Type II Errors: Type I error occurs when the null
    hypothesis is rejected incorrectly, while Type II error occurs when
    the null hypothesis is accepted incorrectly.

-   Statistically different vs clinically/biologically different.
