---
title: "Tests on mean values"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: false
    toc_float: 
       collapsed: true
    #code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```

## Introduction

Hypothesis tests and p-values

-   Motivation
    -   Population mean and sample mean
    -   Understand the question
    -   Explore the data

-   Comparing **one-sample mean** to a standard known mean:

    -   One-Sample t-test (parametric)
    -   One-Sample Wilcoxon Test (non-parametric)

-   Comparing the **means of two independent groups**

    -   Unpaired Two Samples t-test (parametric)
    -   Unpaired Two-Samples Wilcoxon Test (non-parametric)

-   Comparing the **means of paired samples**

    -   Paired Samples t-test (parametric)
    -   Paired Samples Wilcoxon Test (non-parametric)

-   Comparing the **means of more than two groups**

    -   One-Way ANOVA Test (parametric)
    -   Kruskal-Wallis Test (non parametric)

## Prepare data

### **fev** data:

smoking tends to impair lung function arises from studies of pulmonary
function in adults who are long-time smokers. A question then arises as
to whether similar effects can be detected in children who smoke.

Lung function was measured in 654 children seen for a routine checkup in
a particular pediatric clinic. The study subjects were asked whether
they were current smokers.

Forced expiratory volume (FEV) measures the volume of air you can blow
out of your lungs.

```{r}

#dataForAnalysis=read.csv("./data/fev.csv",header=TRUE,row.names=1)
dataForAnalysis <- read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/fev.csv",header=TRUE,row.names=1)

head(dataForAnalysis)

```

## Comparing one-sample mean to a standard known mean

**Question**: Is the mean value of age in this data different than 10?

### Data distribution to decide which test to use

Usually If the sample size is large enough (n \> 30), we can ignore the
distribution of the data and use parametric tests.

```{r}

hist(dataForAnalysis$age)
abline(v=10,lty=2,col="red")

```

### One-Sample t-test (parametric)

```{r}
#T stastic caculation
cohortMean=mean(dataForAnalysis$age)
cohortSD=sd(dataForAnalysis$age)
cohortN=nrow(dataForAnalysis)
populationMean=10

TValue=(cohortMean-populationMean)/(cohortSD/sqrt(cohortN))
TValue

```

```{r}
#one sample t test
t.test(dataForAnalysis$age,mu=populationMean)

```

The result of t.test() function is a list containing the following
components:

-   statistic: the value of the t test statistics

-   p.value: the p-value for the test

-   conf.int: a confidence interval for the mean appropriate to the
    specified alternative hypothesis.

-   estimate: the means of the two groups being compared (in the case of
    independent t test) or difference in means (in the case of paired t
    test).

### One-Sample Wilcoxon Test (non-parametric)

```{r}

wilcox.test(dataForAnalysis$age, mu = populationMean)

```

## Comparing the means of two independent groups

**Question**: Is the mean value of fev between smokers and non-smokers
different?

### Data distribution and summary to decide which test to use

```{r}

hist(dataForAnalysis$fev[which(dataForAnalysis$smoke==1)])
hist(dataForAnalysis$fev[which(dataForAnalysis$smoke==2)])
boxplot(dataForAnalysis$fev~dataForAnalysis$smoke)

tapply(dataForAnalysis$fev,dataForAnalysis$smoke,summary)


```

### Two-Sample t-test (parametric)

T-statistic calculation

```{r}

#T statistic calculation
dataX=(dataForAnalysis$fev[which(dataForAnalysis$smoke==1)])
dataY=(dataForAnalysis$fev[which(dataForAnalysis$smoke==2)])

meanX=mean(dataX)
meanY=mean(dataY)
sdX=sd(dataX)
sdY=sd(dataY)
nX=length(dataX)
nY=length(dataY)

TValue=(meanX-meanY)/(sqrt(sdX^2/nX+sdY^2/nY))
TValue


```

two sample t-test

```{r}

t.test(dataX,dataY,alternative = "two.sided",paired = FALSE)

```

### Unpaired Two-Samples Wilcoxon Test (non-parametric)

```{r}

wilcox.test(dataX,dataY)


```

## Comparing the means of paired samples

## Prepare data

### **sot-covid** data:

The goal of this study was to compare humoral responses to a SARS-CoV-2
vaccine series between solid organ transplant (SOT) recipients and
healthy controls (HCs).

```{r}

#load data
dataForAnalysis=read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/sot-covid.csv",header=TRUE,row.names=1)

head(dataForAnalysis)

```

Let's compare the RBD value between Baseline (rbd1) and three weeks
following first vaccine dose (rbd2*)*.

### Paired Samples t-test (parametric)

Data summary for changes between Baseline and three weeks following
first vaccine dose

```{r}

summary(dataForAnalysis$rbd2-dataForAnalysis$rbd1)

hist(dataForAnalysis$rbd2-dataForAnalysis$rbd1)

plot(dataForAnalysis$rbd1,dataForAnalysis$rbd2)
abline(0,1,lty=2,col="red")

```

T-statistic calculation on "Difference"

```{r,echo=TRUE}
#T stastic caculation
diffValue=dataForAnalysis$rbd1-dataForAnalysis$rbd2

cohortMean=mean(diffValue)
cohortSD=sd(diffValue)
cohortN=length(diffValue)
populationMean=0

TValue=(cohortMean-populationMean)/(cohortSD/sqrt(cohortN))
TValue

```

two sample paired t-test

```{r}

t.test(dataForAnalysis$rbd1,dataForAnalysis$rbd2,paired = TRUE)

```

You will get same results by one sample t-test to compare the difference
with mean=0

```{r}

t.test(dataForAnalysis$rbd1-dataForAnalysis$rbd2,mu=0)


```

### Paired Samples Wilcoxon signed rank test (non-parametric)

```{r}

wilcox.test(dataForAnalysis$rbd1,dataForAnalysis$rbd2,paired = TRUE)


```

## three or more samples ANOVA

The ANOVA test assumes that, the data are normally distributed and the
variance across groups are homogeneous.

Let's compare the Baseline RBD values (rbd1) between different groups of
tissue transplant patients, as well as three weeks following first
vaccine dose (rbd2).

***organ:*** Transplant organ type (0 = healthy control, 1 = kidney, 2 =
liver, 3 = heart, 4 = lung).

### Data visualization

```{r}
boxplot(rbd1   ~ organ, data = dataForAnalysis,main="baseline RBD")

boxplot(rbd2   ~ organ, data = dataForAnalysis,main="RBD at three weeks")

hist(dataForAnalysis$rbd1)
hist(dataForAnalysis$rbd2)


```

### one-way ANOVA

```{r}

dataForAnalysis$organ= factor(dataForAnalysis$organ)

#difference of baseline rbd between organ types
anovaResult1 <- aov(rbd1   ~ organ, data = dataForAnalysis)
summary(anovaResult1)

#difference of rbd at three weeks following first vaccine dose between organ types
anovaResult2 <- aov(rbd2   ~ organ, data = dataForAnalysis)
summary(anovaResult2)


```

### Tukey multiple pairwise-comparisons

```{r}

#TukeyHSD(anovaResult1)
TukeyHSD(anovaResult2)

```

### Kruskal-Wallis rank sum test

Kruskal-Wallis test is a non-parametric alternative to one-way ANOVA
test, which extends the two-samples Wilcoxon test in the situation where
there are more than two groups.

```{r}
#baselilne rbd
kruskal.test(rbd1 ~ organ, data = dataForAnalysis)

#rbd at three weeks following first vaccine dose
kruskal.test(rbd2 ~ organ, data = dataForAnalysis)


```

```{r,warning=FALSE}

#pairwise.wilcox.test(dataForAnalysis$rbd1 , dataForAnalysis$organ,
#                 p.adjust.method = "BH")

pairwise.wilcox.test(dataForAnalysis$rbd2 , dataForAnalysis$organ,
                 p.adjust.method = "BH")

```
