---
title: "Tests on proportions"
author: Shilin Zhao and Leena Choi
output:
  rmdformats::robobook:
    highlight: kate
    toc_depth: 3
    number_sections: no
    toc_float:
      collapsed: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  markdown:
    wrap: 72
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```
## Introduction

Hypothesis tests and p-values

-   Motivation
-   Compare Observed Proportion to an Expected Proportion:
    One-Proportion Z-Test
-   Compare Two Observed Proportions: Two-Proportion Z-Test
-   Compare Multiple Observed Proportions to Expected Proportions:
    Chi-squared test

## Prepare data

**fev** data: Smoking tends to impair lung function. Data in children
who smoke.

**sex**: subject gender (1 = male, 2 = female) **smoke**: smoke smoking
habits (1 = yes, 2 = no)

```{r}

dataForAnalysis <- read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/fev.csv",header=TRUE,row.names=1)
dataForAnalysis$sex=ifelse(dataForAnalysis$sex==1,"male","female")
dataForAnalysis$smoke=ifelse(dataForAnalysis$smoke==1,"yes","no")

head(dataForAnalysis)

```

## Compare Observed Proportion to an Expected Proportion: One-Proportion Z-Test

**Question**: What is the proportion of smokers in this data? Different
than 1/10?

### Data summary

```{r}

table(dataForAnalysis$smoke)/sum(table(dataForAnalysis$smoke))

```

### Statistic value and CI calculation

```{r,echo=TRUE}
#statistic value calculation
totalN=nrow(dataForAnalysis)
observedN=length(which(dataForAnalysis$smoke=="yes"))
observedPropotion=observedN/totalN

## CI of Proportion
observedPropotion-1.96*sqrt(observedPropotion*(1-observedPropotion)/totalN)
observedPropotion+1.96*sqrt(observedPropotion*(1-observedPropotion)/totalN)

expectedPropotion=0.1
ZValue=(observedPropotion-expectedPropotion)/sqrt(expectedPropotion*(1-expectedPropotion)/totalN)
ZValue

```

### One-proportion Z-test

```{r,echo=TRUE}

ztest=prop.test(x = observedN, n = totalN, p = expectedPropotion, correct = FALSE)
ztest
sqrt(ztest$statistic)

```

## Compare Two Observed Proportions: Two-Proportions Z-Test

**Question**: What is the proportion of smokers in male and female
subjects? Different?

### Data summary in two groups

```{r}

t(table(dataForAnalysis$smoke,dataForAnalysis$sex))

t(table(dataForAnalysis$smoke,dataForAnalysis$sex))/(as.vector(table(dataForAnalysis$sex)))


```

### Statistic value calculation

```{r}

nX=length(which(dataForAnalysis$sex=="male"))
nY=length(which(dataForAnalysis$sex=="female"))

nX1=length(which(dataForAnalysis$sex=="male" & dataForAnalysis$smoke=="yes"))
nY1=length(which(dataForAnalysis$sex=="female" & dataForAnalysis$smoke=="yes"))


pX=nX1/nX
pY=nY1/nY
pTotal=(nX1+nY1)/(nX+nY)

ZValue=(pX-pY)/sqrt(pTotal*(1-pTotal)*(1/nX+1/nY))
ZValue

```

### Two-proportion Z-test

```{r}

ztest=prop.test(c(nX1,nY1),c(nX,nY), correct = FALSE)
ztest
sqrt(ztest$statistic)

```

### Same result as Chi-squared test

```{r}

contingencyTable=as.matrix(table(dataForAnalysis$sex,dataForAnalysis$smoke))
chisq.test(contingencyTable,correct = FALSE)

```

## Compare Multiple Observed Proportions to Expected Proportions: Chi-squared test

**Question**: Are the Proportions of transplant tissue types different
between male and female?

### Prepare data

**sot-covid** data

**organ**: Transplant organ type (0 = healthy control, 1 = kidney, 2
=liver, 3 = heart, 4 = lung) **male**: smoke smoking habits (1 = male, 2
= female)

```{r}

#load data
dataForAnalysis=read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/sot-covid.csv",header=TRUE,row.names=1)
head(dataForAnalysis)

```

### Chi-squared test

```{r}
#make contingency table
contingencyTable=as.matrix(table(dataForAnalysis$organ,dataForAnalysis$male))
contingencyTable

chisq.test(contingencyTable)

```
