---
title: "Linear regression"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: false
    toc_float: 
       collapsed: true
    #code_folding: hide
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```

## Introduction

-   Motivation
-   Model assumptions
-   Perform simple and multiple linear regression analysis
-   Model diagnostics


## Data example
We will use the same data, **fev**, used in the lecture '06_TestsOnMeans'. This data have `fev` measuring lung function in children who are smokers or non-smokers. Smoking tends to impair lung function.

* Important variables we use in this lecture:

    - **sex**: subject gender (1 = male, 2 = female)

    - **smoke**: smoke smoking habits (1 = yes, 2 = no)


```{r}

#fev=read.csv("./data/fev.csv",header=TRUE,row.names=1)
fev <- read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/fev.csv",header=TRUE,row.names=1)
fev$sex=ifelse(fev$sex==1,"male","female")
fev$smoke=ifelse(fev$smoke==1,"yes","no")

head(fev)

```

## Simple linear regression

**Question**: Does age associated with fev (lung function )?

### Checking assumptions

**Normality**: Does the dependent variable follows a normal distribution?


```{r}

hist(fev$age, main="", xlab="Age")
hist(fev$fev, main="", xlab="FEV")

```

**Linearity**: Does the relationship between the independent variable and the dependent variable is linear?

```{r}

plot(fev ~ age, data = fev, xlab="Age", ylab="FEV")

```

### Perform a linear regression analysis

Regress `fev` on `age`: $fev \sim b_0 + b_1 age$

```{r}

modelResult <- lm(fev ~ age, data = fev)

summary(modelResult)


```

Interpretation of results:

- The *Coefficients* are estimates of model parameters: the intercept ($b_0$) is 0.431648 and age ($b_1$) is 0.222041. 

    - The estimated regression equation can be written as fev = 0.431648 + 0.222041*age.
    - The intercept ($b_0$) can be interpreted as the predicted fev value at age=0.
    - The coefficient for age ($b_1$) is 0.222041; if age increases by one year, the predicted fev value increases by 0.222041.

- The standard errors (SE) is the standard deviation of sampling distribution of the coefficients. The SE reflects how the coefficient would vary under repeated sampling.

- The R-squared (R2) ranges from 0 to 1 and represents the proportion of variability in fev (dependent variable) explained by the model. For a simple linear regression, R2 is square of the Pearson's correlation coefficient.

- The adjusted R-squared is a modified R-squared to adjust for the number of predictors in the model. It tries to correct for overestimation.

### Model diagnostics

#### Confidence interval of the estimates of model parameters

```{r}

confint(modelResult)

```

If the same study were to be repeated, there is a 95% chance that the interval [0.21, 0.24] contains the true value of $b_1$.

#### Visulization of the fitted line

```{r}

plot(fev$age, fev$fev, main = "", xlab="Age", ylab="FEV")
#abline(modelResult, cex = 1.3, pch = 16, col=2)
lines(fev$age, modelResult$fitted.values, lty=1, col=2)
legend("topleft", "Fitted line", lty=1, col=2)
```

#### Checking the residuals

```{r}

par(mfrow=c(1, 2))
plot(modelResult$fitted.values, modelResult$residuals, main="Residuals vs. Fitted", xlab="Fitted values", ylab="Residuals")
abline(h=0, lty=2)

qqnorm(scale(modelResult$residuals))
abline(0,1,lty=2,col="red")

```


### Predict new data

```{r}

newData <- data.frame(age = c(Subject1 = 5, Subject2 = 10))
predict(modelResult, newData)


```


## Multiple linear regression


Multiple linear regression is an extension of simple linear regression used to predict an outcome based on multiple predictors.

**Question**: Are age and smoking associated with fev (lung function)?


### Data visulization


```{r}

boxplot(fev$fev~fev$smoke, main="", xlab="Smoke", ylab="FEV")

boxplot(fev$age~fev$smoke, main="", xlab="Smoke", ylab="Age")

```

### Perform a linear regression analysis
Regress `fev` on `smoke`: $fev \sim b_0 + b_2 smoke$

```{r}

fit1 <- lm(fev ~ smoke, data = fev)
summary(fit1)

```

### Perform a multiple linear regression analysis
Regress `fev` on `age` and `smoke`: $fev \sim b_0 + b_1 age + b_2 smoke$

```{r}

modelResult <- lm(fev ~ age + smoke, data = fev)

summary(modelResult)
confint(modelResult)

```


Interpretation of results:

- The estimated regression line equation can be written as fev = 0.367373 + 0.230605 * age -0.208995 * Smoker

- How to interpret the coefficients?

### Effects of variables

```{r, message = FALSE}

library(effects)
plot(allEffects(modelResult))


```





