---
title: "Linear regression"
author: Shilin Zhao and Leena Choi
output:
  rmdformats::robobook:
    highlight: kate
    toc_depth: 3
    number_sections: no
    toc_float:
      collapsed: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  markdown:
    wrap: 72
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```
## Introduction

-   Motivation
-   Model assumptions
-   Perform simple and multiple linear regression analysis
-   Model diagnostics

## Data example

We will use the same data, **fev**, used in the lecture
'06_TestsOnMeans'. This data have `fev` measuring lung function in
children who are smokers or non-smokers. Smoking tends to impair lung
function.

-   Important variables we use in this lecture:

    -   **sex**: subject gender (1 = male, 2 = female)

    -   **smoke**: smoke smoking habits (1 = yes, 2 = no)

```{r}

#fev=read.csv("./data/fev.csv",header=TRUE,row.names=1)
fev <- read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/fev.csv",header=TRUE,row.names=1)
fev$sex=ifelse(fev$sex==1,"male","female")
fev$smoke=ifelse(fev$smoke==1,"yes","no")

head(fev)

```

## Simple linear regression

**Question**: Does age associated with fev (lung function )?

### Checking assumptions

**Normality**: Does the dependent variable follows a normal
distribution?

```{r}

hist(fev$age, main="", xlab="Age")
hist(fev$fev, main="", xlab="FEV")

```

**Linearity**: Does the relationship between the independent variable
and the dependent variable is linear?

```{r}

plot(fev ~ age, data = fev, xlab="Age", ylab="FEV")

```

### Perform a linear regression analysis

Regress `fev` on `age`: $fev \sim b_0 + b_1 age$

```{r}

modelResult <- lm(fev ~ age, data = fev)

summary(modelResult)


```

Interpretation of results:

-   The *Coefficients* are estimates of model parameters: the intercept
    ($b_0$) is 0.431648 and age ($b_1$) is 0.222041.

    -   The estimated regression equation can be written as fev =
        0.431648 + 0.222041\*age.
    -   The intercept ($b_0$) can be interpreted as the predicted fev
        value at age=0.
    -   The coefficient for age ($b_1$) is 0.222041; if age increases by
        one year, the predicted fev value increases by 0.222041.

-   The standard errors (SE) is the standard deviation of sampling
    distribution of the coefficients. The SE reflects how the
    coefficient would vary under repeated sampling.

-   The R-squared (R2) ranges from 0 to 1 and represents the proportion
    of variability in fev (dependent variable) explained by the model.
    For a simple linear regression, R2 is square of the Pearson's
    correlation coefficient.

-   The adjusted R-squared is a modified R-squared to adjust for the
    number of predictors in the model. It tries to correct for
    overestimation.

### Model diagnostics

#### Confidence interval of the estimates of model parameters

```{r}

confint(modelResult)

```

If the same study were to be repeated, there is a 95% chance that the
interval [0.21, 0.24] contains the true value of $b_1$.

#### Visualization of the fitted line

```{r}

plot(fev$age, fev$fev, main = "", xlab="Age", ylab="FEV")
#abline(modelResult, cex = 1.3, pch = 16, col=2)
lines(fev$age, modelResult$fitted.values, lty=1, col=2)
legend("topleft", "Fitted line", lty=1, col=2)
```

#### Checking the residuals

```{r}

par(mfrow=c(1, 2))
plot(modelResult$fitted.values, modelResult$residuals, main="Residuals vs. Fitted", xlab="Fitted values", ylab="Residuals")
abline(h=0, lty=2)

qqnorm(scale(modelResult$residuals))
abline(0,1,lty=2,col="red")

```

### Predict new data

```{r}

newData <- data.frame(age = c(Subject1 = 5, Subject2 = 10))
predict(modelResult, newData)


```

### Visualization of confidence interval

```{r}
preds <- predict(modelResult, data = fev, interval = 'confidence')

plot(fev$age,fev$fev,pch=16)

#add fitted regression line
abline(modelResult,col=2)
#add dashed lines for confidence bands
lines(fev$age, preds[ ,3], lty = 'dashed', col = 'blue')
lines(fev$age, preds[ ,2], lty = 'dashed', col = 'blue')

```

## Multiple linear regression

Multiple linear regression is an extension of simple linear regression
used to predict an outcome based on multiple predictors.

**Question**: Are age, gender, and smoking associated with fev (lung
function)?

### Data visualization

```{r}

boxplot(fev$fev~fev$smoke, main="", xlab="Smoke", ylab="FEV")
boxplot(fev$fev~fev$sex, main="", xlab="Smoke", ylab="FEV")


```

### Perform a linear regression analysis

Regress `fev` on `smoke`: $fev \sim b_0 + b_1 smoke$

```{r}

fit1 <- lm(fev ~ smoke, data = fev)
summary(fit1)
```

Regress `fev` on `sex`: $fev \sim b_0 + b_1 sex$

```{r}
fit2 <- lm(fev ~ sex, data = fev)
summary(fit2)
```

### Visualization of age, gender, and smoking together

#### Age (by smoking) vs FEV

```{r}

plot(fev$age, fev$fev, main = "", xlab="Age", ylab="FEV",col=factor(fev$smoke),pch=16)
legend("topleft",legend=levels(factor(fev$smoke)),bty="n",col=1:2,pch=16)

```

#### Age (by gender) vs FEV

```{r}
plot(fev$age, fev$fev, main = "", xlab="Age", ylab="FEV",col=factor(fev$sex),pch=16)
legend("topleft",legend=levels(factor(fev$sex)),bty="n",col=1:2,pch=16)

```

### Perform a multiple linear regression analysis

Regress `fev` on `age` and `smoke`:
$fev \sim b_0 + b_1 age + b_2 smoke + b_3 sex$

```{r}

modelResult <- lm(fev ~ age + smoke+sex, data = fev)

summary(modelResult)
confint(modelResult)

```

Interpretation of results:

-   The estimated regression line equation can be written as fev =
    0.24 + 0.23 \* age -0.15 \* Smoker -0.31 \* Male

-   How to interpret the coefficients?

```{r}

plot(fev$age, fev$fev, main = "", xlab="Age", ylab="FEV",col=factor(fev$smoke),pch=16)
legend("topleft",legend=levels(factor(fev$smoke)),bty="n",col=1:2,pch=16)
abline(modelResult$coefficients["(Intercept)"],modelResult$coefficients["age"],lwd=2)
abline(modelResult$coefficients["(Intercept)"]+modelResult$coefficients["smokeyes"],modelResult$coefficients["age"],col=2,lwd=2)
legend("topleft",legend=levels(factor(fev$smoke)),bty="n",col=1:2,pch=16)

```

```{r}

plot(fev$age, fev$fev, main = "", xlab="Age", ylab="FEV",col=factor(fev$sex),pch=16)
legend("topleft",legend=levels(factor(fev$sex)),bty="n",col=1:2,pch=16)
abline(modelResult$coefficients["(Intercept)"],modelResult$coefficients["age"],lwd=2)
abline(modelResult$coefficients["(Intercept)"]+modelResult$coefficients["sexmale"],modelResult$coefficients["age"],col=2,lwd=2)
legend("topleft",legend=levels(factor(fev$sex)),bty="n",col=1:2,pch=16)

```

### Effects of variables

```{r, message = FALSE}

library(effects)
plot(allEffects(modelResult))


```
