---
title: "Logistic Regression"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: false
    toc_float: 
       collapsed: true
    #code_folding: hide
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```
## Introduction

-   Data example
-   Simple logistic regression analysis
-   Multiple logistic regression analysis

## Data example
We will use the same data, **fev**, used in the lecture '06_TestsOnMeans'. This data have `fev` measuring lung function in children who are smokers or non-smokers. Smoking tends to impair lung function.

* Important variables we use in this lecture:

    - **sex**: subject gender (1 = male, 2 = female)
    - **smoke**: smoke smoking habits (1 = yes, 2 = no)
    - **age**: subject age at time of measurement (years)
    - **fev**: measured FEV (liters per second)
    - **height**: subject height at time of measurement (inches)

```{r}

#fev=read.csv("./data/fev.csv",header=TRUE,row.names=1)
fev <- read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/fev.csv",header=TRUE,row.names=1)
fev$sex=ifelse(fev$sex==1,"male","female")
fev$smoke=factor(ifelse(fev$smoke==1,"yes","no"))

head(fev)

```

## Simple logistic regression

**Question**: Does smoking associated with age?

### Logistic regression assumptions

-   The outcome is a binary or dichotomous variable: e.g., yes vs. no, positive vs. negative, 1 vs. 0.
-   The observations are independent.
-   There is a linear relationship between the logit of $\mbox{Pr}(Y=1 | X)$ and predictors: $\mbox{logit}\left\{\mbox{Pr}(Y=1 | X)\right\} = X \beta$.

### Data visulization

```{r}

table(fev$smoke)
hist(fev$age, main="", xlab="Age")

boxplot(fev$age ~ fev$smoke, main="", xlab="Smoking status", ylab="Age")

```

### Perform logistic regression analysis

$\mbox{logit}\left\{\mbox{Pr}(y=1 | X)\right\} = b_0 + b_1  age$, where $y = 1$ if smoking is "yes"; otherwise $y = 0$.

```{r}

modelResult <- glm(smoke ~ age, data = fev, family = "binomial")
summary(modelResult)

```

Interpretation of results:

- The *Coefficients* are estimates of model parameters: the intercept ($b_0$) is -7.74391 and age ($b_1$) is 0.48364.

- The estimated regression equation can be written as follow: $\mbox{log}\left (\frac{Pr(y=1 | X)}{1-Pr(y=1 | X)}\right) = -7.74391 + 0.48364 \; \mbox{age}$

- $Pr(y=1 | X)$ = $\frac{e^{(-7.74391  \: +  \: 0.48364 \: age)}}{1+e^{(-7.74391 \: +  \: 0.48364 \: age)}}$,  where $Pr(y=1 | X)$ is the probability of being a smoker at a specific age.

### Predict new data

```{r}

newData <- data.frame(age = c(Subject1=10,Subject2=50))
predictionResult=predict.glm(modelResult, newData,type ="response") #probabilities
predictionResult

```

### Odds ratio

- Let $p = Pr(y=1 | X)$. The $\frac{p}{1-p}$ is called odds, or odds of success, which is the ratio of "successes" to "non-successes". It reflects the likelihood that the event will occur. $\mbox{logit} \left \{ Pr(y=1 | X) \right \} = \mbox{logit}(p) = \mbox{log}\left (\frac{p}{1-p}\right)$ is called log odds.  

- Odds ratio (OR) is the ratio of two odds:
$$\frac{\left (p_1/(1-p_1)\right)}{\left (p_2/(1-p_2)\right)}$$
- It represents the ratio of the odds that an event will occur (for example, being a smoker) in one scenario (for example, age = 2), $p_1$, compared to the odds of the event occurring (being a smoker) in another scenario (for example, age = 1), $p_2$. The OR in a logistic regression measures the association between the outcome variable (being smoker) and a predictor variable (age).

- Based on the equation from the model, log($\frac{p}{1-p}$) = -7.74391 + 0.48364 x age:
    - We can estimate the OR being a smoker at Age1 (e.g., age = 2) compared to Age2 (e.g., age = 1):
    
    $e^{(-7.74391 + 0.48364*Age1)}/e^{(-7.74391 + 0.48364*Age2)}$ = $e^{(-7.74391 + 0.48364*Age1)-(7.74391 + 0.48364*Age2)}$ = $e^{0.48364*(Age1-Age2)}$

    - We typically report the OR by taking exponentiation of the coefficient. $e^{0.48364}=1.62$. It indicates that with one year of age increase, the odds of being a smoker will increase by a factor of 1.62.


```{r}

#odds ratio
exp(coefficients(modelResult))

#CI for odds ratio
exp(confint(modelResult))

```

## Multiple logistic regression

**Question**: Does smoking associated with age, height, and sex?

### Data visulization

```{r}

table(fev$sex,fev$smoke)
plot(fev[,c("age","height")], xlab="Age", ylab="Height")

boxplot(fev$height~fev$smoke, xlab="Smoking status", ylab="Height")


```

### Perform logistic regression analysis

```{r}

modelResult <- glm(smoke ~ age + height + sex, data = fev, family = "binomial")
summary(modelResult)

```

### Effects of variables

```{r, message=FALSE}

library(effects)
plot(allEffects(modelResult), axes=list(x=list(rug=FALSE)))


```
