---
title: "Describing Data Exercises"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r, echo = FALSE, message = FALSE}
library(Hmisc)
```

# Load data

Download `https://hbiostat.org/data/repo/titanic3.sav` and load it into R. This file is in a binary format that R can understand. Find a function that can load a file with a saved R object.

```{r}
```

  * Did it load?
  * What are the data types? Try `str(<data>)` where "<data>" is the variable name of the data set.

```{r}
```

# Set factor variables

Modify the "survived" column, by changing it into a factor variable. It should have two levels, "no" and "yes".

```{r}
```

# Describe the data

Use a function to describe this data set.

```{r}
```

  * What stands out?

# Create a table one

What would make a good exposure variable? What other variables should you include?

```{r}
```

  * What stands out?

# Visually inspect the data

A density plot might be a good choice for a continuous variable. Do we have any?

Can you think of another type of plot that might be interesting to see (if we knew how to create it)?

```{r}
```

  * What stands out?
