---
title: "Visualizing Data Exercises"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r, echo = FALSE, message = FALSE}
library(ggplot2)
```

# Load data

Download `https://hbiostat.org/data/repo/titanic3.sav` and load it into R. Turn "survived" into a factor variable.

```{r}
```

# Single variable

If plotting a single variable, which plot/geom would you select for the following variables?
* pclass
* survived
* sex
* age
* fare
* embarked

Produce these six plots using `ggplot` (not `qplot`).

```{r}
```

# ECDF

Create a ECDF plot with "age".

```{r}
```

About 3/4(ths) of people were at or under what age?

# Relationship with "survived"

Repeat the plots by faceting over "survived" (don't show the plot for "survived").

```{r}
```

# Relationship between class and fare

Create a plot showing the relationship between "pclass" and "fare".

```{r}
```

# Continuous data with "survived"

Using "survived" as the x-axis, produce plots for each of the continuous variables.

```{r}
```

# Continuous/continuous

Now produce a plot with the two continuous variables.

```{r}
```

Add one of the discrete variables as a color.

```{r}
```

# Adding model output

We'll create a simple model that predicts survival using sex, age, and fare. The new variable "xSurv" will be a probability between 0 and 1.

```{r}
tdat <- titanic3[complete.cases(titanic3[, c('survived','sex','age','fare')]),]
m <- glm(survived ~ sex + age + fare, data = tdat, family = binomial)
tdat[,'xSurv'] <- predict(m, type = 'response')
```

Create a plot showing the relationship between "age" and "xSurv". Also include the "survived" variable.

```{r}
```
