---
title: "Base Graphics Exercises"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Load data

Download `https://hbiostat.org/data/repo/titanic3.sav` and load it into R. Turn "survived" into a factor variable.

```{r}
```

# Single variable

If plotting a single variable, which plot would you select for the following variables?

* pclass
* survived
* sex
* age
* fare
* embarked

Produce these six plots.

```{r}
```

# ECDF

Create a ECDF plot with "age". Call the `plot` function on the `ecdf` function of the variable of interest.

Change the x-axis label and the main plot title.

```{r}
```

# Relationship with "survived"

Create a barplot showing the relationship between "sex" and "survived".

```{r}
```

# Relationship between class and fare

Create a boxplot showing the relationship between "pclass" and "fare". When `boxplot` uses two variables, it expects the variables to be provided in the formula syntax. This syntax looks like "Y ~ X" where "Y" is the continuous variable and "X" is the categorical variable.

```{r}
```

# Continuous/continuous

Create a scatterplot of "age" and "fare".

```{r}
```

Add "sex" as a color.

```{r}
```
