---
title: "R Programming Exercises"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Load data

Download `https://hbiostat.org/data/repo/titanic3.sav` and load it into R. Turn "survived" into a factor variable.

```{r}
```

Show the rows with the five youngest passengers.

```{r}
```

  * How many 1st class passengers from Southampton survived?
  * How many 2nd class passengers from Cherbourg had a missing "fare"?
  * How many 3rd class passengers from Queenstown were women?

```{r}
```

Use a "for" loop to find the mean "fare" for "embarked". Remove any missing values for "fare".

```{r}
```

Use one of the apply functions to find the mean "fare" for "embarked". Remove any missing values for "fare".

```{r}
```

Create a subset of adult men (age >= 18). Then restrict to rows where "sibsp" equals 1 and "ticket" is not missing.

```{r}
```

Create a subset of adult women (age >= 18). Then restrict to rows where "sibsp" equals 1 and "ticket" is not missing.

```{r}
```

Merge the men and women subsets by "ticket" and "cabin".

```{r}
```

Create a new subset (of the merged data) requiring the string "Mrs" in the woman's name column. We will assume that this data set contains spouse pairs.

```{r}
```

From the spouse pairs:

  * How many times does both the husband and wife survive?
  * Neither?
  * Just the wife?
  * Just the husband?

```{r}
```

  * What is the average age difference?
  * How about absolute difference?

```{r}
```

Show the rows with the biggest age gap (both positive and negative).

```{r}
```
