---
title: "Tests on mean values"
output:
  rmdformats::robobook:
    highlight: kate
    toc_depth: 3
    number_sections: no
    author: Shilin Zhao and Leena Choi
    toc_float:
      collapsed: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  markdown:
    wrap: 72
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```
```{r setup,echo=FALSE}

knitr::opts_chunk$set(echo = FALSE,eval=FALSE)


```

## Exercise

-   Comparing one-sample mean to a standard known mean

    -   One-Sample t-test (parametric)
    -   One-Sample Wilcoxon Test (non-parametric)

-   Comparing the means of two independent groups

    -   Unpaired Two Samples t-test (parametric)
    -   Unpaired Two-Samples Wilcoxon Test (non-parametric)

-   Comparing the means of paired samples

    -   Paired Samples t-test (parametric)
    -   Paired Samples Wilcoxon Test (non-parametric)

-   Comparing the means of more than two groups

    -   One-Way ANOVA Test (parametric)
    -   Kruskal-Wallis Test (non parametric)

## Prepare data

```{r,message=FALSE,echo=TRUE}

#Get Titanic data
library(Hmisc)
getHdata(titanic3)
head(titanic3)

dataForAnalysis=titanic3

```

## Comparing one-sample mean to a standard known mean

**Question**: What's the mean age in titanic data? Is it different than
30?

### Exercise 1: Show **Mean and SD** of age

```{r}


```

### Exercise 2: Plot **distribution** of age

```{r}

```

### Exercise 3: Is the mean age different than 30, using **one-sample t test**

```{r}

```

### Exercise 4: Is the mean age different than 30, using **Wilcoxon rank sum test**

```{r}


```

## Comparing the means of two independent groups

**Question**: Is there any age difference between survivors and
non-survivors?

### Exercise 1: Testing by **two-sample t-test**

```{r}


```

### Exercise 2: Testing by **Wilcoxon rank sum test**

```{r}


```

## Comparing the means of three or more independent groups

**Question**: Is there any age difference between people in different
ticket class (column name: pclass)?

### Exercise 1: Data visualization of age distribution by **boxplot**

```{r}


```

### Exercise 2: Testing by **ANOVA**

```{r}



```

### Exercise 3: Which two groups are different? By **Tukey multiple pairwise-comparisons**

```{r}



```

### Exercise 4: Testing by **Kruskal-Wallis test**

```{r}



```

### Exercise 5: Which two groups are different? By **pairwise Wilcoxon rank sum test**

```{r}



```
