---
title: "Tests on mean values"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: false
    toc_float: 
       collapsed: true
    #code_folding: hide
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```



```{r setup,echo=FALSE}

knitr::opts_chunk$set(echo = FALSE,eval=FALSE)


```



## Exercise

-   Comparing one-sample mean to a standard known mean

    -   One-Sample T-test (parametric)
    -   One-Sample Wilcoxon Test (non-parametric)

-   Comparing the means of two independent groups

    -   Unpaired Two Samples T-test (parametric)
    -   Unpaired Two-Samples Wilcoxon Test (non-parametric)

-   Comparing the means of paired samples

    -   Paired Samples T-test (parametric)
    -   Paired Samples Wilcoxon Test (non-parametric)

-   Comparing the means of more than two groups

    -   One-Way ANOVA Test (parametric)
    -   Kruskal-Wallis Test (non parametric)

## Prepare data

```{r,message=FALSE,echo=TRUE}

#Get Titanic data
library(Hmisc)
getHdata(titanic3)
head(titanic3)

dataForAnalysis=titanic3

```

## Comparing one-sample mean to a standard known mean

**Question**: What's the mean age in titanic data? Is it different than 30?

### Exercise 1: Show **Mean and SD** of age

```{r}

mean(dataForAnalysis$age,na.rm=TRUE)
sd(dataForAnalysis$age,na.rm=TRUE)

```

### Exercise 2: Plot **distribution** of age

```{r}

hist(dataForAnalysis$age)

```

### Exercise 3: Is the mean age different than 30, using **one-sample t test**

```{r}
#one sample t test
t.test(dataForAnalysis$age,mu=30)

```

### Exercise 4: Is the mean age different than 30, using **Wilcoxon rank sum test**

```{r}

wilcox.test(dataForAnalysis$age,30)

```




## Comparing the means of two independent groups

**Question**: Is there any age difference between suviviors and non-suviviors?


### Exercise 1: Testing by **two-sample t-test**

```{r}

t.test(dataForAnalysis$age[which(dataForAnalysis$survived==0)],dataForAnalysis$age[which(dataForAnalysis$survived==1)])

```

### Exercise 2: Testing by **Wilcoxon rank sum test**

```{r}

wilcox.test(dataForAnalysis$age[which(dataForAnalysis$survived==0)],dataForAnalysis$age[which(dataForAnalysis$survived==1)])

```



## Comparing the means of three or more independent groups

**Question**: Is there any age difference between people in different ticket class (column name: pclass)?


### Exercise 1: Data visualization of age distribution by **boxplot**

```{r}

boxplot(age   ~ pclass, data = dataForAnalysis)

```



### Exercise 2: Testing by **ANOVA**

```{r}

anovaResult <- aov(age   ~ pclass, data = dataForAnalysis)
summary(anovaResult)

```


### Exercise 3: Which two groups are different? By **Tukey multiple pairwise-comparisons**

```{r}

TukeyHSD(anovaResult)


```



### Exercise 4: Testing by **Wilcoxon rank sum test**

```{r}

kruskal.test(age  ~ pclass, data = dataForAnalysis)


```



