---
title: "Tests on proportions"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: false
    toc_float: 
       collapsed: true
    #code_folding: hide
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
body{
  font-size: 14pt;
}
</style>
```



```{r setup,echo=FALSE}

knitr::opts_chunk$set(echo = FALSE,eval=FALSE)


```



## Exercise

-   Compare Observed Proportion to an Expected Proportion: One-Proportion Z-Test
-   Compare Two Observed Proportions: Two Proportions Z-Test
-   Compare Multiple Observed Proportions to Expected Proportions: Chi-square test


```{r,message=FALSE,echo=TRUE}

#Get Titanic data
library(Hmisc)
getHdata(titanic3)
head(titanic3)

dataForAnalysis=titanic3

```


## Compare Observed Proportion to an Expected Proportion: One-Proportion Z-Test

*Question:* What's the proportion of survivors in Titanic data, different than 0.4?  

### Exercise 1: show count and proportion of survivors                          

```{r}

table(dataForAnalysis$survived)
table(dataForAnalysis$survived)/sum(table(dataForAnalysis$survived))

```


### Exercise 2: Using Z test to compare the observed propotion with 0.4                        

```{r}

prop.test(x = length(which(dataForAnalysis$survived==1)), n = sum(table(dataForAnalysis$survived)), p = 0.4, correct = FALSE)

```


## Compare Two Observed Proportions: Two Proportions Z-Test

*Question:* What's the proportion of survivors in male and female, are they different?

### Exercise 1: show count and proportion of survivors in male and female                         

```{r}

table(dataForAnalysis$sex,dataForAnalysis$survived)
table(dataForAnalysis$sex,dataForAnalysis$survived)/as.vector(table(dataForAnalysis$sex))

```


### Exercise 2: Using Z test to compare the two propotions

```{r}

nX=length(which(dataForAnalysis$sex=="male"))
nY=length(which(dataForAnalysis$sex=="female"))

nX1=length(which(dataForAnalysis$sex=="male" & dataForAnalysis$survived==1))
nY1=length(which(dataForAnalysis$sex=="female" & dataForAnalysis$survived==1))

prop.test(c(nX1,nY1),c(nX,nY), correct = FALSE)


```


## Compare Multiple Observed Proportions to Expected Proportions: Chi-square test

*Question:* What's the proportion of survivors in people in different class (pclass), are they different?

### Exercise 1: show count and proportion of survivors in different class                     

```{r}

table(dataForAnalysis$pclass,dataForAnalysis$survived)
table(dataForAnalysis$pclass,dataForAnalysis$survived)/as.vector(table(dataForAnalysis$pclass))

```

### Exercise 2: Using Chi-square test to compare the three propotions

```{r}

chisq.test(table(dataForAnalysis$pclass,dataForAnalysis$survived))

```








