---
title: "Linear regression"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: false
    toc_float: 
       collapsed: true
    #code_folding: hide
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
</style>
```

```{r setup,echo=FALSE}

knitr::opts_chunk$set(echo = FALSE,eval=FALSE)


```



## Exercise

-   Data assumptions: show data distribution and discuss about Normality and Linearity
-   Perform the linear regression analysis
-   Model evaluation: discuss model result


## Data

```{r,echo=TRUE}

#load data
dataForAnalysis=read.csv("./data/sot-covid.csv",header=TRUE,row.names=1)
dataForAnalysis$organ=factor(dataForAnalysis$organ)

head(dataForAnalysis)

```

**Question**: Is RBD at three weeks (rbd2) associated with basedline RBD (rbd1) and other factors including organ and age?

### Exercise 1: Data visulization for  rbd2, rbd1, age, and organ

```{r}

plot(dataForAnalysis[,c("rbd2","rbd1","age")])

boxplot(dataForAnalysis$rbd2~dataForAnalysis$organ)

```


### Exercise 2: Perform the linear regression analysis, rbd2~rbd1+age+organ

```{r}

modelResult <- lm(rbd2 ~ rbd1+age+organ, data = dataForAnalysis)

```


### Exercise 3: Model interpretation, show coefficients, confidence interval, and p values 

```{r}

summary(modelResult)
confint(modelResult)

```



### Exercise 4: USe the model to predict the rbd2 value for a new subject.


```{r,echo=TRUE,eval=TRUE}

newData <- data.frame(age = c(Subject1=3,Subject2=4),rbd1=c(0.426,0.153),organ=c(1,2))


```




