---
title: "Linear regression"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: false
    toc_float: 
       collapsed: true
    #code_folding: hide
---

```{=html}
<style>
h2 {
  background-color: #D4DAEC;
}
</style>
```
```{r setup,echo=FALSE}

knitr::opts_chunk$set(echo = FALSE,eval=FALSE)


```

## Exercise

-   Data assumptions: show data distribution and discuss about Normality and Linearity
-   Perform the linear regression analysis
-   Model evaluation: discuss model result

## Data

```{r,echo=TRUE}

#load data
dataForAnalysis <- read.csv("https://raw.githubusercontent.com/couthcommander/sip_computing/main/labs/sot-covid.csv",header=TRUE,row.names=1)

dataForAnalysis$organ=factor(dataForAnalysis$organ)

head(dataForAnalysis)

```

**Question**: Is ELISA response to nucleocapsid protein at three weeks (nc2) associated with baseline (nc1) and other factors including organ and age?

### Exercise 1: Data visualization

Use some pictures (histogram, scatter plot, or box plot,...) to show the distribution of all factors (nc2, nc1, age, and organ); Think about:

(1) Assess the data assumptions (Normality and Linearity) for linear regression;

(2) Based on (1), do you think we need to perform a log transformation on nc1 and nc2?

(3) What is the relationships between nc1, nc2, age, and organ;

```{r}


```

### Exercise 2: Perform simple linear regression analysis

Using nc2 as Y, nc1 as X

```{r}


```

### Exercise 3: Model interpretation

(1) Is nc1 significantly associated with nc2? What is the coefficient of nc1? What does it mean?
(2) Use a scatter plot to show the nc1 in X axis, and nc2 in Y axis. Add the fitted regression line.
(3) Make residual plot to assess the Homoscedasticity assumption.

```{r}


```

### Exercise 4: Perform multiple linear regression analysis

Using nc2 as Y; nc1, age, and organ as X

```{r}


```

### Exercise 5: Use the model to predict the nc2 value for a new subject.

```{r}


```
